---
title: "Qualitative Activity Recognition of Weight Lifting Exercises"
author: "Ayako Nagao"
date: "10/31/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE)
```

### 1.Overview  
The goal of this project is to build the model to classify 5 different ways of weight lifting exercises

### 2.Data  

```{r loadData,cache=TRUE}
url <- 'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv'
data <- read.csv(url)
dim(data)
```

```{r splitData}
library(caret)
set.seed(10103)
# use 70% for training, 30% for testing
inTrain <- createDataPartition(y=data$classe,p=0.7,list=FALSE)
train <- data[inTrain,]
test <- data[-inTrain,]
obs <- dim(train)[1]; var <- dim(train)[2]
```
The training dataset contains `r obs` observations and `r var` features. In this dataset there are many features with NA or empty strings. All features with NA, around 98% of the observations are NA(missing), all features with empty strings, also around 98% of the observations are empty strings(missing). Thus we omit those features and also omit features irrelevant to activity recognition such as user names, timestamp, and new window.

```{r summaryTraining}
# # count NAs for each column
# NAcounts <- colSums(is.na(train))
# unique(NAcounts)
# # count empty strings "" in character class columns
# dataClass <- sapply(train,class)
# chrVar <- names(dataClass[dataClass=='character'])
# numVar <- names(dataClass[dataClass!='character'])
# emptyCounts <- colSums(train[,chr]=='')
# unique(emptyCounts)
# # check near zero variance variables among not NA numeric variables
# nonNAnumeric <- intersect(numVar,names(NAcounts[NAcounts==0]))
# nzv <- nearZeroVar(train[,nonNAnumeric],saveMetrics=TRUE)$nzv
# nzv
# data.frame(Data.class=c('numeric/integer','character'),
#            Counts=c(length(numVar),length(chrVar)),
#            NA.or.Empty=c(sum(NAcounts!=0),sum(emptyCounts!=0)))
# 
# 
# 
# 


# xyz <- names(train)[grepl('[xyz]$',names(train))]
# notXyz <- names(train[grepl('[^xyz]$',names(train))])
# length(xyz)
# length(notXyz)


```
### 3.Explorattory Analysis  
```{r featureExtraction}

```

### 4.Modeling  

```{r modeling}

```
### 5.Results  
